\documentclass[phd,tocprelim]{cornell}

%
% tocprelim option must be included to put the roman numeral pages in the
% table of contents
%
% The cornellheadings option will make headings completely consistent with
% guidelines.
%
% This sample document was originally provided by Blake Jacquot, and
% fixed up by Andrew Myers.
%
%Some possible packages to include
\usepackage{amsmath}
\usepackage{graphicx,pstricks}
\usepackage{graphics}
\usepackage{moreverb}
\usepackage{subfigure}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{hangcaption}
\usepackage{txfonts}
\usepackage{listings}
\usepackage{palatino}
\usepackage{algorithm}
\usepackage{multirow}
\usepackage{url}
\usepackage{bm}
% \usepackage{siunitx}
\usepackage{xspace}
\usepackage{tabularx}

\usepackage{algcompatible}
\usepackage[floatrow]{trivfloat}
\trivfloat{algorithm}
\renewcommand\algorithmname{ALGORITHM}

%\let\oldAA\AA
%\renewcommand{\AA}{\text{\normalfont\oldAA}}

\newcommand\nd{\textsuperscript{nd}\xspace}
\newcommand\rd{\textsuperscript{rd}\xspace}
\newcommand\nth{\textsuperscript{th}\xspace} %\th is taken already

\newcommand{\angstrom}{\text{\normalfont\AA}}

\lstset{basicstyle=\small\ttfamily,breaklines=true}
% \usepackage{ctex}

%if you're having problems with overfull boxes, you may need to increase
%the tolerance to 9999
\tolerance=9999

\bibliographystyle{plain}
%\bibliographystyle{IEEEbib}

\renewcommand{\caption}[1]{\singlespacing\hangcaption{#1}\normalspacing}
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.75}

\def\beq{\begin{eqnarray}}
\def\eeq{\end{eqnarray}}
\def\V{\mathcal{V}}
\def\C{\mathcal{C}}
\def\P{\mathcal{P}}

\def\NMCdiff{{N_{\mathrm{MC}}^{\mathrm{uniq}}}}
\def\Nd{{N_d}}
\def\Nddiff{{N_{d}^{\mathrm{uniq}}}}

\def\vecD{\vec{D}}
\def\veca{\vec{\alpha}}
\def\vecb{\vec{\beta}}
\def\ia{i_\alpha}
\def\ja{j_\alpha}
\def\ib{i_\beta}
\def\veciD{\vec{i}_D}
\def\vecia{\vec{i}_\alpha}
\def\vecib{\vec{i}_\beta}
\def\veciDa{\vec{i}_{D\alpha}}
\def\veciDb{\vec{i}_{D\beta}}
\def\veciaa{\vec{i}_{\alpha\alpha}}
\def\veciba{\vec{i}_{\beta\alpha}}
\def\veciab{\vec{i}_{\alpha\beta}}
\def\vecibb{\vec{i}_{\beta\beta}}

\title {Fast Semistochastic Heat-Bath Configuration Interaction}
\author {Junhao Li}
\conferraldate {August}{2019}
\degreefield {Ph.D.}
\copyrightholder{Junhao Li}
\copyrightyear{2019}

\begin{document}

\maketitle
\makecopyright

\begin{abstract}
In this thesis, I present my work on the fast semistochastic heatbath configuration interaction (Fast SHCI), which is an efficient algorithm for electronic structure calculations on large clusters.

Fast SHCI is more than an order of magnitude faster than other algorithms in its category (selected configuration interaction plus perturbation theory), and also much faster than other essentially exact algorithms for many chemical systems.
This thesis provides an in-depth description of the Fast SHCI algorithm and its implementation. I use SHCI to compute electronic structures of several chemical systems and the homogeneous electron gas.
Some of these calculations are more accurate than those achieved by other high-order quantum chemistry methods.  Others treat systems larger than those that can be treated by other equally accurate methods.

My implementation of SHCI uses a modular design, which not only makes the library highly extensible but also contributes several generic distributed computing building blocks to the open-source community. 
In this thesis, I also describe my design and implementation of these generic components.

Finally, I also provide a brief discussion of the usability of general software engineering best practices for the development of medium-scale scientific software packages with lessons learned from designing, developing, and leading the development of our SHCI package.
Medium-scale scientific software packages are common in scientific research where a small group of researchers works on the same code base.
Due to the differences in the requirements, some best practices that are common in industry need to be adjusted to be useful for these projects.

\end{abstract}

\begin{biosketch}
Junhao Li was born and grew up in Shanghai, China.
From a young age, he had a strong interest in science and engineering, and enjoyed disassembling all kinds of home appliances.

Junhao attended Shanghai Jiao Tong University from 2009, graduating with a B.S. in physics and a B.S.E in computer science in 2013.
While an undergraduate, he did research in various fields, including semiconductor fabrication, photovoltaics, finite element analysis of electromagnetic field, density functional theory, and social networks.

In the fall of 2013, Junhao came to Cornell University to persue a Ph.D. in physics.
During his Ph.D., he mainly worked with professor Cyrus Umrigar on the development and application of highly accurate quantum chemistry methods.
He developed the fast heatbath configuration interaction method, which is more than an order of magnitude faster than other methods in its category, and achieved significantly higher accuracy than other well developed methods, such as DMRG and FCIQMC, on many systems.
he also did research in molecular dynamics and defects with professor James Sethna.

After obtaining a master's degree at Cornell, Junhao spent a summer as an intern at Google in California in 2016.
He then returned to Ithaca, brought back the engineering merits he learned from Google, and developed a highly efficient and extensible quantum chemistry package, Arrow, as well as several open-sourced generic high performance computing libraries.
\end{biosketch}

\begin{dedication}
This document is dedicated to all Cornell graduate students.
\end{dedication}

\begin{acknowledgements}
First of all, I would like to thank my advisor Professor Cyrus Umrigar.
His commitment to physics research is exemplary and inspiring.
He led me into the field of quantum chemistry, which is an indispensable foundation of highly accurate quantum simulations.
All my knowledge of quantum Monte Carlo are inherited from him, and my work on heatbath configuration interaction is impossible without his guidance and support.
He taught me valuable research skills, and helped me go through the process of publishing my first first-author paper, on a top academic journal.
I am also extremely grateful to him for always being supportive in the decisions I made, the ideas I wanted to try, and being tolerant and direct when I made mistakes.
He made my six years of Ph.D. life a truly rewarding experience.

I would also like to thank Professor James Sethna.
It was a pleasure to work with Jim on defects and molecular dynamics.
His enthusiasm and optimism inspires me whenever I face chanllenging problems.
I also learned a lot from his tremendous insights on physics and data analysis.

I also thank all the graduate students I worked with, especially Adam Holms, Matt Otten, and Matt Bierbaum.
They helped me getting started in my research, patiently answered all my questions, and gave me valuable advices on my projects.
I could not imagine how much longer it would take me to get start and how many detours I might have taken without their help.

I would like to give special thanks to my parents for giving me birth and raising me up.
They gave me a warm family as I grew up, and consistently supported me as I persue my degree at Cornell.
Thanks also to my elementary school, middle school, high school, and undergraduate teachers and friends back in China for all the invaluable lessons I learned from them, the wonderful memories I had with them, and helps and encouragements I received from them.

Finally, I acknowledge the financial support from Cornell University physics department, National Science Foundation and the Air Force Office of Scientific Research, and the computing resources support from Pittsburg Computing Center, the Argonne National Lab, NERSC, and Google Cloud.
Thank you and I hope you are proud.
\end{acknowledgements}

\contentspage
\tablelistpage
\figurelistpage

\normalspacing \setcounter{page}{1} \pagenumbering{arabic}
\pagestyle{cornell} \addtolength{\parskip}{0.5\baselineskip}

\chapter{Introduction}
\label{ch:intro}
A major problem in the electronic structure theory is solving the many-body Schr\"{o}dinger's equation accurately and efficiently, and there is usually a trade-off between accuracy and efficiency.
Density functional theory (DFT)~\cite{ParYan-BOOK-89,DreGro-BOOK-90,kohn1999nobel} is the most popular method and it is very efficient,
but it uses approximate density functionals, and thus it is often not accurate enough.
%In addition, the choice of pseudopotentials relies a lot on empirical experiences.
Coupled cluster with single, double, and perturbative triple excitations CCSD(T)~\cite{raghavachari1989fifth} is often considered as the gold standard of quantum chemistry.
It is very accurate for single reference systems where a single determinant in the many-body wavefunction has a large amplitude, such as many organic molecules, but not for multireference or strongly-correlated systems.
Density matrix renormalization group (DMRG)~\cite{white1993density,white1999ab,chan2002highly,chan2011density,ShaCha-JCP-12,olivares2015ab,schollwock2005density,GuoLiCha-JCTC-18}
and full configuration interaction quantum Monte Carlo (FCIQMC)~\cite{BooThoAla-JCP-09,CleBooAla-JCP-10,PetHolChaNigUmr-PRL-12,BooGruKreAla-Nat-13,HolChaUmr-JCTC-16}
are
systematically improvable and can give exact solutions given infinite computing resources, but rapidly get expensive with the number of electrons and the size of the basis set.

Fast heatbath configuration interaction (SHCI)~\cite{HolTubUmr-JCTC-16,ShaHolJeaAlaUmr-JCTC-17,HolUmrSha-JCP-17,SmiMusHolSha-JCTC-17,MusSha-JCTC-17,ChiHolOttUmrShaZim-JPCA-18,LiOttHolShaUmr-JCP-18}
is another systematically improvable method capable of providing essentially exact energies for small chemical systems.
In common with FCIQMC, the computational cost of the method scales exponentially in the number of electrons, but SHCI is much faster than FCIQMC.
The comparison with DMRG is more involved.
SHCI is much faster than DMRG for small moderately correlated systems.
SHCI scales exponentially with system size with a prefactor that is typically small, but the prefactor grows with
the strength of the correlation.
DMRG scales exponentially with the $(D-1)/D$-th power of the system size (where D is the system dimension) with a prefactor that is typically larger,
but the prefactor is not very sensitive to the strength of the correlation.

SHCI is an example of the selected configuration interaction plus perturbation theory (SCI+PT)
methods~\cite{HurMalRan-JCP-73,BuePey-TCA-74,EvaDauMal-CP-83,CimPer-JCoP-87,Har-JCP-91,BytRue-CP-09,KelPerBarGre-JCP-14,CoeMurPat-CPL-14,Eva-JCP-14,SceAppGinCaf-JCoC-16,GarSceLooCaf-JCP-17,LooSceBloGarCafJac-JCP-18,TubLevHaiHeaWha-ARX-18}.
SCI+PT methods have two stages.
In the first stage, a variational wavefunction is constructed iteratively, starting from
a determinant that is expected to have a significant amplitude in the final wavefunction, e.g., the Hartree-Fock determinant.
Each iteration of the variational stage has three steps: the selection of important determinants, construction of the Hamiltonian matrix, and
iterative diagonalization of the Hamiltonian matrix.
In the second stage, 2$^{\rm nd}$-order perturbation theory is used to improve upon the variational energy.

The Fast SHCI algorithm has greatly improved the efficiency of both stages of SCI+PT calculations.
First, it greatly speeds up the determinant selection and Hamiltonian construction by using the heat-bath selection criterion and the fast Hamiltonian construction algorithm, and, second, it drastically reduces the central processing unit (CPU) cost as well as the memory cost of performing the perturbation step by using a 3-step perturbation algorithm.
%The Fast SHCI algorithm~\cite{LiOttHolShaUmr-JCP-18} has greatly improved the efficiency of both stages of SHCI~\cite{HolTubUmr-JCTC-16,ShaHolJeaAlaUmr-JCTC-17,HolUmrSha-JCP-17,SmiMusHolSha-JCTC-17,MusSha-JCTC-17,ChiHolOttUmrShaZim-JPCA-18}.
%First, it greatly speeds up the determinant selection and Hamiltonian construction by using the heatbath selection criterion and the fast Hamiltonian construction algorithm, and, second, it drastically reduces the central processing unit (CPU) cost as well as the memory cost of performing the perturbation step by using a 3-step perturbation algorithm.
With fast SHCI, we can employ two orders of magnitude larger variational wavefunctions than other SCI+PT methods and can incorporate the entire Hilbert space in the perturbative correction calculation.

I apply fast SHCI to the homogeneous electron gas (HEG) system.
HEG is one of the most fundamental models in condensed-matter physics and is commonly used for benchmarking new methods.
With a tunable parameter controlling the density of the electrons, HEG provides a wide range of interacting fermion systems from weakly correlated to strongly correlated.
HEG is also the foundation of density functional theory (DFT).
In this thesis, I focus on the HEG systems in the mid to high-density region.
In this region, the basis set incompleteness error is the dominating error for most quantum chemistry methods.
SHCI works well with large basis sets and thus naturally solves the basis set incompleteness problem.
In our calculation, we use up to $4\times10^4$ orbitals, which is over an order of magnitude more the number of orbitals used in previous calculations with other methods, such as FCIQMC.

I also apply fast SHCI to the chromium dimer system.
Chromium dimer is a challenging strongly-correlated system that has been used as a benchmark molecule for a variety of methods~\cite{Scu-JCP-91,KurYan-JCP-11,PurZhaKra-JCP-15,MaManOlsGag-JCTC-16,VanMalVer-JCTC-16,GuoWatHuSunCha-JCTC-16}.
I calculate both the energy at the Cr$_2$ equilibrium geometry and its entire potential energy curve.
For the equilibrium geometry, with 2 billion variational determinants, we achieve extremely high accuracy, which beats the accuracy of well-developed methods, such as the density matrix renormalization group (DMRG).
For the potential energy curve, we include in our Hilbert space up to 28 active electrons with a cc-pVDZ basis.
This Hilbert space is several orders of magnitude larger than the Hilbert space used in other systematically improvable methods.

One of the core components of our SHCI software is a generic cluster computing library, which consists of a high-performance implementation of MapReduce and several parallel data containers.
MapReduce~\cite{dean2008mapreduce,dean2010mapreduce} provides users a high-level abstraction for defining their parallel computation and takes care of the intricate low-level execution steps internally.
Each MapReduce operation consists of two phases: a map phase where each input is mapped to a set of intermediate key/value pairs, and a reduce phase where the pairs with the same key are reduced to a single key/value pair according to a user-specified reduce function.
Many algorithms are expressible with this model, including the algorithm for finding important determinants and evaluating the perturbative correction in our SHCI method.
MapReduce implementations can automatically allocate resources and execute these algorithms in parallel.
Most industrial implementations focus on handling big data rather than achieving higher performance.
Our library focuses on high performance, and to achieve better performance, we introduce three main improvements to the MapReduce algorithm: eager reduction, fast serialization, and special treatment for a small fixed key range.
Our library achieves an order of magnitude faster performance than popular cluster computing library, Spark, on several compute-intensive MapReduce tasks.

The usage of software engineering techniques is also essential to the successful development and maintenance of the SHCI software.
Software engineering studies the management of the software development process in a scientific and systematic way.
It helps to make the software development faster, easier and making the software more robust and more extensible.
Software engineering is especially important to scientific software development because if the software has poor quality, the results may be unreliable, which can lead to wrong conclusions and mislead future research.
There are many software engineering practices in the industry.
I exam the most common ones and discuss their applicability in the setting of scientific software development.

\input{docs/shci.tex}
\input{docs/heg.tex}
\input{docs/cr2.tex}
\input{docs/benchmark.tex}
\input{docs/blaze.tex}
\input{docs/eng.tex}

\chapter{Conclusion}
\label{ch:con}

Fast semistochastic heatbath configuration interaction (Fast SHCI) is an efficient algorithm for electronic structure calculations.
It is more than an order of magnitude faster than other selected configuration interaction plus perturbation theory algorithms, and also much faster than other essentially exact quantum chemistry methods, such as DMRG or FCIQMC, in many cases.

The key reasons that fast SHCI is fast are:
\begin{itemize}
    \item The heatbath criterion is easy to evaluate, and important determinants that meet this criterion can be found efficiently by using precomputed double excitation lists.
    \item The fast Hamiltonian construction algorithm significantly speeds up the Hamiltonian construction process by using helper arrays.
    \item The 3-step perturbation algorithm speeds up the perturbation correction calculation regardless of the memory of the system.
\end{itemize}

SHCI achieves accurate results on homogeneous electron gas in the mid to high-density region due to its high efficiency and ability to work with large basis sets.
SHCI also successfully produces accurate results on huge Hilbert spaces of Chromium dimers.

Modular design makes the SHCI code easy to maintain and extend, and also produces several generic distributed computing building blocks, which makes writing high-performance parallel software much easier than using raw MPI routines directly.

Finally, regarding scientific software engineering, it is important to see that scientific software is becoming more and more complex, and thus it is crucial to keep quality of the code high so that the scientific results from this software are reliable and the process of extending the software in order to experiment new scientific ideas is efficient.
We can borrow the experience from industrial software engineering, but due to the differences between industrial and scientific research, some common practices which are generally believed to be beneficial to industrial software engineering will have to be adjusted to really benefit scientific software development.

\bibliography{sampleThesis,umrigar,toulouse,chan,needs,zhang,chemistry,realistic}

\end{document}
